One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2020-12-07 paper-v2.tex
change one sentence in iterated inla description
change back to chicago style biblography????
Add project file to help avoid spurious whitespace differences.
fix typo in linearisation equation
update N post figure ref

The half-normal detection function depends on the strictly positive parameter $\sigma^2$.  We choose a log link and consider $g(s | \phi)$, where $\log\sigma^2 = \phi$.  However, this cannot be estimated in standard INLA formulation as the predictor is not linear in its parameters.  To address this we use an approximate method based on a first-order Taylor approximation of the non-linear predictor.  Let $\eta(\bm{u}) = \log \tilde{\lambda}(\bm{u})$ where $\bm{u}$ is the parameter vector of the predictor (which includes $\phi$ as well as any parameters for fixed and random effects).  Denote by $\bar{\eta}_{\bm{u}_0}$ the first-order Taylor approximation of $\eta(\bm{u})$ at the point $\bm{u}_0$.  That is
\begin{equation*}
\bar{\eta}_{\bm{u}_0} = \eta(\bm{u_0}) + \bm{B}_{\bm{u}_0}(\bm{u} - \bm{u}_0)
\end{equation*}
where $\bm{B}_{\bm{u}_0}$ is the derivative matrix of $\eta(\bm{u})$ evaluated at $\bm{u}_0$.  Since $\bar{\eta}_{\bm{u}_0}$ is linear in $\bm{u}$ it is possible to fit an approximate model in INLA by replacing the non-linear $\eta$ with the linear Taylor approximation $\bar{\eta}_{\bm{u_0}}$.   The question then is how to choose the linearisation point $\bm{u}_0$.  The approach taken is via a fixed point iteration scheme.  There are many possible choices here and we describe the approach in the current implementation of \texttt{inlabru} \citep{bachl_inlabru_2019}.  However, more options are likely to be added in the future.  We denote by $\bar{p}_{\bm{v}}$ the posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$ and define the functional
\begin{equation}
	f(\bar{p}_{\bm{v}}) = \left\{ \argmax_{\bm{u}_i} \: \bar{p}_{\bm{v}}(u_i | \bm{Y}), \quad i = 1, \ldots, n_{\bm{u}} \right\}
\end{equation}  
where $n_{\bm{u}}$ is the length of $\bm{u}$.  Each $\bar{p}_{\bm{v}} (u_i | \bm{Y})$ is approximated and the mode estimated during the model fitting using the standard INLA method.
We then seek a fixed point of this functional so that $\bm{u}_0 = f(\bar{p}_{\bm{u}_0})$.  This fixed point is identified via an iterative scheme.  Given a current choice of linearisation point $\bm{v}$, estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_v)$ to use as the new linearisation point.  Iterate this procedure until a fixed point is identified within a chosen tolerance.

@@

We plot the posterior abundance estimate using a Monte Carlo method.
\begin{figure}[h]
	\begin{center}
		\includegraphics[scale=0.525]{figures/realized_abundance_posterior.png}
		\caption{Posterior abundance}
		\label{fig:realized-abundance-posterior}
	\end{center}
\end{figure}
Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.  Integrating the mean of the posterior intensity field provides a point estimate for the abundance, i.e. the expected abundance.  We approximate the posterior $\pi(N | \bm{Y})$ by a Monte Carlo method.  Taking $m$ Monte Carlo samples  $\lambda^{(1)}, \ldots, \lambda^{(m)}$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)})$. Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$. \autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.  This allows us to estimate the probability for any specific value of realised abundance $n$ by estimated $\pi(N = n | \lambda)$.  Strictly speaking

@@

The quantile maps also suffer from a potential difficulty in interpretation. The temptation is to perceive the maps as showing possible intensity surfaces that could have produced the observed data.  However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for all prediction values to simultaneously achieve the 0.025 or 0.975 quantile.  A similar problem will occur if mapping the marginal probability of exceeding a threshold at each prediction location.  We believe these caveats make the map difficult to use for most non-statistically trained audiences.  To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile map and 12,400 for the 0.975 quantile map.  A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance. However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).  This demonstrates that the tendency to interpret these maps as possible intensities leads to interpretations that are inconsistent with the very model that generated the maps.

@@

\clearpage
\bibliographystyle{humannat}
\bibliography{paper}

>>>>

The half-normal detection function depends on the strictly positive parameter $\sigma^2$.  We choose a log link and consider $g(s | \phi)$, where $\log\sigma^2 = \phi$.  However, this model cannot be directly fitted using INLA since $\log g(s | \phi)$ is not linear in its parameters.  To address this we use an approximate method based on a first-order Taylor approximation of the non-linear predictor.  Let $\eta(\bm{u}) = \log \tilde{\lambda}(\bm{u})$ where $\bm{u}$ is the parameter vector of the predictor (which includes $\phi$ as well as any parameters for fixed and random effects).  Denote by $\bar{\eta}_{\bm{u}_0}$ the first-order Taylor approximation of $\eta(\bm{u})$ at the point $\bm{u}_0$.  That is
\begin{equation*}
\bar{\eta}_{\bm{u}_0} (\bm{u}) = \eta(\bm{u_0}) + \bm{B}_{\bm{u}_0}(\bm{u} - \bm{u}_0)
\end{equation*}
where $\bm{B}_{\bm{u}_0}$ is the derivative matrix of $\eta(\bm{u})$ evaluated at $\bm{u}_0$.  Since $\bar{\eta}_{\bm{u}_0}$ is linear in $\bm{u}$ it is possible to fit an approximate model in INLA by replacing the non-linear $\eta$ with the linear Taylor approximation $\bar{\eta}_{\bm{u_0}}$.   The question then is how to choose the linearisation point $\bm{u}_0$.  The approach taken is via a fixed point iteration scheme.  There are many possible choices here and we describe the approach in the current implementation of \texttt{inlabru} \citep{bachl_inlabru_2019}.  However, more options are likely to be added in the future.  We denote by $\bar{p}_{\bm{v}}$ the posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$ and define the functional
\begin{equation}
	f(\bar{p}_{\bm{v}}) = \left\{ \argmax_{\bm{u}_i} \: \bar{p}_{\bm{v}}(u_i | \bm{Y}), \quad i = 1, \ldots, n_{\bm{u}} \right\}
\end{equation}
where $n_{\bm{u}}$ is the length of $\bm{u}$.  Each $\bar{p}_{\bm{v}} (u_i | \bm{Y})$ is approximated and the mode estimated during the model fitting using the standard INLA method.
We then seek a fixed point of this functional so that $\bm{u}_0 = f(\bar{p}_{\bm{u}_0})$.  This fixed point is identified via an iterative scheme.  Given a current choice of linearisation point $\bm{v}$, estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_v)$ to use as the new linearisation point.  Iterate this procedure until a fixed point is identified within a chosen tolerance.

@@

We plot the posterior abundance estimate using a Monte Carlo method.
\begin{figure}[h]
	\begin{center}
		\includegraphics[scale=0.525]{figures/N_posterior.png}
		\caption{Posterior abundance.  Grey area marks two Monte-carlo standard errors above and below the mean}
		\label{fig:realized-abundance-posterior}
	\end{center}
\end{figure}
Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.  Integrating the mean of the posterior intensity field provides a point estimate for the abundance, i.e. the expected abundance.  We approximate the posterior $\pi(N | \bm{Y})$ by a Monte Carlo method.  Taking $m$ Monte Carlo samples  $\lambda^{(1)}, \ldots, \lambda^{(m)}$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)})$. Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$. \autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.  This allows us to estimate the probability for any specific value of realised abundance $n$ by estimated $\pi(N = n | \lambda)$.  Strictly speaking

@@

The quantile maps also suffer from a potential difficulty in interpretation. The temptation is to perceive the maps as showing possible intensity surfaces that could have produced the observed data.  However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for all prediction values to simultaneously achieve the 0.025 or 0.975 quantile.  A similar problem will occur if mapping the marginal probability of exceeding a threshold at each prediction location.  We believe these caveats make the map difficult to use for most non-statistically trained audiences.  To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 2,800 for the 0.025 quantile map and 10,200 for the 0.975 quantile map.  A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance. However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).  This demonstrates that the tendency to interpret these maps as possible intensities leads to interpretations that are inconsistent with the model that generated the maps.

@@

\clearpage
\bibliographystyle{chicago}
\bibliography{paper}
